Synthesizing Quotient Lenses

### 2018-03-12 papers/QuotientLenses/main.tex
change the bibtex example

Our solution to this shortcoming is to introduce {\em Quotient Regular
Expressions} (QREs), which enable programmers to simultaneously specify a
regular expression and a set of canonical representatives for matching
strings.  Given an QRE, we can automatically infer the corresponding
canonization functions.  
\ksf{Define comma\_name here as a really simple example? }
\afm{Right now comma\_name is just a regular expression.  Further details in Fig
2}

@@

let bib_to_end : (lens in bibtex $\Leftrightarrow$ end_note) = preamble . label . author_lens . title_lens
\end{lstlisting}
\caption{An example bijective lens \cd{bib_to_end} between data
matching \cd{bibtex} and \cd{end_note} regular 
expressions, which are omitted for brevity.  
\iflater
\bcp{And a readability suggestion:
  define a named constant for a quote character.}\fi}
\label{fig:example-lens}
\end{figure}

@@

\subsection{Specifying \bibtex{} Using QREs}
\label{subsec:qre-expressions}
\ksf{Are these QRE combinators or QRE lens combinators?  
We are currently mushy about this distinction.  
The section needs a pass to clarify this point.
The headers of this section and the next imply that this section is
about QREs and the next section is about QRE lenses, but this section
is where we introduce the QRE lens combinators, right?  If so, we need
to do something to make the structure clearer.}

@@

\begin{lstlisting}
let p_wsp = project [ \n\t\r]$^+$ $\to$ " "
\end{lstlisting}
\iflastminute
\bcp{Can we display blanks as a distinct ``visible blank'' symbol
(throughout)?  That will make examples easier to read.}\bcp{I tried turning
on the showstringspaces option at the top, but that doesn't do everything we
want--it only works inside quotes.  But the latex ``textvisiblespace'' command
produces a single one, and the lstlisting ``escapechar'' option will allow
us to use it inside listings.  But we should save this for the final
tweaking stage.}
% Here is an example: if you set escapechar=\%, you could use it inside the
% listings as follows: \begin{lstlisting} t %\leftarrow% 0 \end{lstlisting}
\fi

@@

Another QRE primitive is the sequential composition combinator,
written ``;''.
For an example of its use, suppose we have already defined a QRE
\codefont{comma_to_space} that accepts names in either space-separated
or name-separated format and chooses the space-separated format as canonical.
Suppose we have also already defined a
QRE \codefont{project_in_space}, which projects away extra white space
between the first and last name in the space-separated name format. 
We can use the sequential composition combinator to combine these two QREs:
\begin{lstlisting}
let name = comma_to_space ; space_name_p
\end{lstlisting}

The permutation QRE combinator, \kw{perm}, allows data to be unordered. For example,
the following instance of \kw{perm} allows label, author and title fields
(which we assume have been defined earlier) to appear in any order.
%
\begin{lstlisting}
let bib_fields = perm (label, bib_author, bib_title)
\end{lstlisting}
%
To normalize the field separators, one can specify 
in an optional \kw{with} clause that the components of the
permutation are conjoined by another QRE. For instance, below, we normalize
whitespace between fields, leaving only a single newline.
%
\begin{lstlisting}
let bib_fields = perm (label, bib_author, bib_title) with (project ("," .wsp+) $\to$ ",\n")
\end{lstlisting}

@@

For instance, assume that $f(s)=$ `` '' for all whitespace strings $s$. Then the
project QRE \kw{p\_wsp} defined above can
be expressed using the normalize combinator as
\lstinline{normalize (wsp+, " ", f)}.

Figure~\ref{fig:example-qre} gives the complete QRE specification for the
\bibtex{}.
\begin{figure}
\begin{lstlisting}
let wsp = [ \n\t\r]

(* define name representations with a space and with a comma *)
let space_name = first_name . wsp+ . last_name
let comma_name = last_name . "," . wsp+ . first_name

(* squash QRE maps comma_name to space_name *)
let comma_to_space = squash comma_name $\to$ space_name using c_to_s

(* map arbitrary whitespace into " " as a canonical representative *)
let p_wsp = project wsp+ $\to$ " "

(* Project away whitespace in space_name representation *)
let space_name_p = first_name . p_wsp . last_name

(* Compose the two QREs to get a canonical representation for a name *)
let name = comma_to_space ; space_name_p

(* define rest of bibtex fields *)
let name_sep = p_wsp . "and" . p_wsp
let bib_names = name . (name_sep . name)$^*$)
let bib_author = "author = \"" . bib_names . "\""
let title = word . (p_wsp . word)$^*$
let bib_title = "title = {" . title . "}"

(* allow any permutation of fields interspersed with arbitrary whitespace *)
let bib_fields = perm (label, bib_author, bib_title) with (project ("," . wsp+) $\to$ ",\n")
let bibtex = "@book{" . bib_fields . "}"
\end{lstlisting}
\caption{QRE specification of \bibtex{} records.\ksf{In what sense is this a QRE
spec vs a QRE lens spec?}\afm{I don't really like the way this is set up.  The
use of the sequential composition operator makes the QREs appear really
operational, and in my opinion makes them seem less cool, and distinct from
canonizers, than they actually are.  I include a proposed alternative in the
next figure.  However, this would mean that the previous section which describes
the sequential composition operator would require it be talked about without the
concrete example.}}
\label{fig:example-qre}
\end{figure}

\begin{figure}
\begin{lstlisting}
let wsp = [ \n\t\r]+
let wsp_s = project wsp $\to$ " "

(* define name representations with a space and with a comma *)
let space_name = first_name . wsp_s . last_name
let comma_name = last_name . "," . wsp_s . first_name

@@

(* define rest of bibtex fields *)
let bib_names = name . (wsp_s . "and" . wsp_s . name)$^*$
let bib_author = "author = \"" . bib_names . "\""
let title = word . (wsp_s . word)$^*$
let bib_title = "title = {" . title . "}"

(* allow any permutation of fields interspersed with arbitrary whitespace *)
let bib_fields = perm (label, bib_author, bib_title) with (project ("," . wsp) $\to$ ",\n")
let bibtex = "@book{" . bib_fields . "}"
\end{lstlisting}
\caption{QRE specification of \bibtex{} records.\ksf{In what sense is this a QRE
spec vs a QRE lens spec?}}
\label{fig:example-qre-anders-proposal}
\end{figure}

>>>>

Our solution to this shortcoming is to introduce {\em Quotient Regular
Expressions} (QREs), which enable programmers to simultaneously specify a
regular expression and a set of canonical representatives for matching
strings.  Given an QRE, we can automatically infer the corresponding
canonization functions.  As an example, consider the following QRE for writing
author names:
%
\begin{lstlisting}
let wsp_sp = project wsp $\to$ " "
let comma_name = first_name . "," . wsp_sp . last_name
\end{lstlisting}
%
In this example, \lstinline{wsp} is an existing regular expression for a
nonempty sequence of whitespace characters, and \lstinline{first_name} and
\lstinline{last_name} are existing regular expressions for first and last names.
The QRE \lstinline{wsp_sp} is a QRE with the same underlying language as
\lstinline{wsp}, but with a single canonical representative: \lstinline{" "}.
This is then used to build up \lstinline{comma_name}, a QRE with an
underlying language of two names, separated by a command and whitespace, where
the names with a single space between them are canonical.

@@

let bib_to_end : (lens in bibtex $\Leftrightarrow$ end_note) = preamble . label . author_lens . title_lens
\end{lstlisting}
\caption{An example bijective lens \cd{bib_to_end} between data
matching \cd{bibtex} and \cd{end_note} regular 
expressions, which are omitted for brevity.  
\ifdraft
\bcp{And a readability suggestion:
  define a named constant for a quote character.}
\fi}
\label{fig:example-lens}
\end{figure}

@@

\subsection{Specifying \bibtex{} Using QREs}
\label{subsec:qre-expressions}
\ksf{Are these QRE combinators or QRE lens combinators?  
We are currently mushy about this distinction.  
The section needs a pass to clarify this point.
The headers of this section and the next imply that this section is
about QREs and the next section is about QRE lenses, but this section
is where we introduce the QRE lens combinators, right?  If so, we need
to do something to make the structure clearer.}\afm{I believe this are QRE
combinators, so we should be good.}

@@

\begin{lstlisting}
let wsp_sp = project wsp $\to$ " "
\end{lstlisting}
\iflastminute
\bcp{Can we display blanks as a distinct ``visible blank'' symbol
(throughout)?  That will make examples easier to read.}\bcp{I tried turning
on the showstringspaces option at the top, but that doesn't do everything we
want--it only works inside quotes.  But the latex ``textvisiblespace'' command
produces a single one, and the lstlisting ``escapechar'' option will allow
us to use it inside listings.  But we should save this for the final
tweaking stage.}
% Here is an example: if you set escapechar=\%, you could use it inside the
% listings as follows: \begin{lstlisting} t %\leftarrow% 0 \end{lstlisting}
\fi

@@

The permutation QRE combinator, \kw{perm}, allows data to be unordered. For example,
the following instance of \kw{perm} allows label, author and title fields
(which we assume have been defined earlier) to appear in any order.
%
\begin{lstlisting}
let bib_fields = perm (label, bib_author, bib_title)
\end{lstlisting}
%
To normalize the field separators, one can specify 
in an optional \kw{with} clause that the components of the
permutation are conjoined by another QRE. For instance, below, we normalize
whitespace between fields, leaving only a single newline.
%
\begin{lstlisting}
let bib_fields = perm (label, bib_author, bib_title) with (project ("," . wsp) $\to$ ",\n")
\end{lstlisting}

Another QRE primitive is the sequential composition combinator, written ``;''.
For an example of its use, suppose we have already defined a QRE,
\lstinline{canonized_whitespace}, that accepts XML documents and chooses
documents with no whitespace as canonical. Suppose that we also have defined a
QRE, \lstinline{canonized_order}, which accepts whitespace-normalized XML
documents, and chooses a specific ordering of XML elements as canonical. We can
use the sequential composition combinator to combine these two QREs into
\lstinline{canonized_whitespace ; canonized_order}, a QRE that accepts all XML
documents, and chooses ordered XML documents without whitespace as canonical.

@@

For instance, assume that $f(s)=$ `` '' for all whitespace strings $s$. Then the
project QRE \lstinline{wsp_sp} defined above can
be expressed using the normalize combinator as
\lstinline{normalize (wsp," ", f)}.

Figure~\ref{fig:example-qre} gives the complete QRE specification for the
\bibtex{}.
\begin{figure}
\begin{lstlisting}
let wsp = [ \n\t\r]+
let wsp_sp = project wsp $\to$ " "
let last_name = [A-Z][a-z]
let first_name = [A-Z][a-z]

(* define name representations with a space and with a comma *)
let space_name = first_name . wsp_sp . last_name
let comma_name = last_name . "," . wsp_sp . first_name

@@

(* define rest of bibtex fields *)
let bib_names = name . (wsp_sp . "and" . wsp_sp . name)$^*$
let bib_author = "author = \"" . bib_names . "\""
let title = word . (wsp_sp . word)$^*$
let bib_title = "title = {" . title . "}"

(* allow any permutation of fields interspersed with arbitrary whitespace *)
let bib_fields = perm (label, bib_author, bib_title) with (project ("," . wsp) $\to$ ",\n")
let bibtex = "@book{" . bib_fields . "}"
\end{lstlisting}
\caption{QRE specification of \bibtex{} records.\ksf{In what sense is this a QRE
spec vs a QRE lens spec?}}
\label{fig:example-qre}
\end{figure}
